---
title: "Computations with `dplyr::mutate`"
author: "Silvie Cinková "
date: "2025-08-07"
echo: true
shift-heading-level-by: -1
format:
  revealjs:
    plugins: [notes]
    toc: true
    slide-number: true
    notes: true
    pdf-export: true
    css: atrium.css
    chalkboard: false
    scrollable: true
    code-block-height: 650px
    echo: true
    output-location: default
    preview-links: auto
  pdf:
    toc: true
    number-sections: true
    include-in-header: atrium.tex
  html:
    toc: true
    number-sections: true
    css: atrium.css
    embed-resources: true
---

## Libraries and data

```{r}
library(gapminder)
library(dplyr)
library(magrittr)
americas_2002 <- gapminder %>% filter(continent == "Americas", year == 2002)
gapminder_2002 <- gapminder %>% filter(year == 2002) %>% select(!c(year))
```

## `mutate` {.smaller}

-   works in two contexts:

    -   adds a new column

    -   edits an already existing column

-   can add/edit several columns at once

-   uses its dedicated helper functions

    -   `across(), c_across()`

-   one more grouping function: `rowwise()`

## Adding a new column with `mutate` {.smaller}

```{r}
americas_2002 %>% 
  mutate(GDP_millions = pop * gdpPercap/1000000 ) %>% 
  slice(1:10)

```

## Adding a new column

```{r}
americas_2002 %>% 
  mutate(richest = max(gdpPercap), 
         perc_richest = round(gdpPercap/richest, 2)) 
 
  
```

::: notes
Notice how you can concatenate new columns in one `mutate` action and that the second uses the result of the first within the same command. But I am not sure whether this is always granted.
:::

## Adding a new column computed on groups

```{r}
gapminder_2002 %>%
  group_by(continent) %>% 
  mutate(richest = max(gdpPercap), 
         perc_richest = round(gdpPercap/richest, 2)) %>% 
  ungroup()
```

::: notes
Here we do the same operation as with Americas, but this time for each continent. The `richest` column has identical values for countries on one continent.

Note the difference in the behavior of `max()` and division (`/`). The `max` function always looks at the entire group whereas division operates on each row independently.
:::

## Adding a value on a condition: `if_else`

```{r}
gapminder_2002 %>%
  mutate(is_Asia = if_else(
    condition = continent == "Asia", 
    true = "Asian country",
    false = "Not in Asia"))
```

## Adding a value on a condition: `case_when` {.smaller}

```{r}
gapminder %>% filter(year == 2002) %>% 
  mutate(is_Asia = case_when(
    continent == "Asia" ~  "Asian country",
    continent == "Africa" ~ "African country",
    continent == "Europe" ~ "European country",
    continent == "Americas" ~ "American country",
    continent == "Oceania" ~ "Oceanian country",
    .default = NA
    ))
```

::: notes
Note the specific syntax of `case_when`!
:::

## `mutate` an existing column with `across`

-   convert pop to millions and round to 3 decimal points

```{r}
gapminder_2002 %>%
  mutate(across(pop, ~ round(.x/1000000, 3)))
```

::: notes
`across` lists the columns on which you want to perform the same transformation. Note its peculiar syntax: it is inside `mutate` and incorporates both the enumeration of the columns to transform and the transformation, which is written in the formula notation with tilde and `.x` for the first argument.
:::

## `mutate` an existing column on a selection of columns

```{r}
gapminder_2002 %>% select(!c(continent)) %>%
  mutate(across(
    where(~ is.numeric(.x)),
          ~ round(.x, digits = 0)
    )    )
```

::: notes
`across` works well with all column-selecting helper functions, such as `where` or `starts_with`.
:::

## Compute summary stats across columns {.smaller}

-   `sum` , `mean` and many other functions collapse all values in a column.

-   Override with `rowwise() %>% mutate(c_across(….)`

```{r}
iris %>% slice(1:3) %>% rowwise() %>%
  mutate(sum_leafmeasure = sum(c_across(where(~ is.numeric(.x))))) %>%   ungroup()
```

## Without `rowwise` and `c_across` 😐 {.smaller}

```{r}
iris %>% slice(1:3) %>% 
  mutate(sum_leafmeasure = sum(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)) 
```

## The same with `mean`

```{r}
iris %>% slice(1:3) %>% rowwise() %>%
  mutate(mean_leafmeasure = mean(c_across(!c(Species)))) %>%   ungroup()
```

## Common stat. functions requiring `rowwise` {.smaller}

-   `sum()` -- sums values across columns

-   `mean()` -- computes average across columns

-   `sd()` -- standard deviation

-   `var()` -- variance

-   `min()` / `max()` -- minimum / maximum

-   `median()` -- median value

-   `IQR()` -- interquartile range
