---
title: "Read someone else's code"
author: "Silvie Cinková "
date: 2025-08-9
echo: true
shift-heading-level-by: -1
format:
  revealjs:
    plugins: [notes]
    toc: true
    slide-number: true
    notes: true
    pdf-export: true
    css: atrium.css
    chalkboard: false
    scrollable: true
    code-block-height: 650px
    echo: true
    #output-location: slide
    preview-links: auto
  pdf:
    shift-heading-level-by: -2
    number-sections: true
    include-in-header: atrium.tex
    geometry: portrait,margin=1cm
    highlight-style: tango
  html:
    toc: true
    number-sections: true
    css: atrium.css
    embed-resources: true
---

```{r}
library(tidyverse)
```

```{r}
billionaires_df <- read_tsv("../datasets_ATRIUM/billionaires_combined.tsv")
billionaires_2020 <- billionaires_df %>% 
  filter(time == 2020) %>% select(daily_income, person, world_6region)
```

### The task performed by the code

The data set compiles charts of world's billionaires between 2002 and 2020. This means that the same person can occur more than one time, but their details may differ. Plot billionaires in each world region (column `world_6region`). Use boxplots to plot the distribution of `daily_income` among all billionaires in the given world region and text labels with values from the column `person` for outliers. You may have to plot each region separately, that is, not break one plot into facets but run the/a plotting script individually for each world region.

## Solution


```{r}
#| fig-width: 20
#| fig.height: 20
library(ggrepel)
billionaires_outliers <- billionaires_2020 %>% 
  group_by(world_6region) %>%
  mutate(Q3 = quantile(daily_income, 0.75), 
         IQR_col = IQR(daily_income), 
         outliers_above = Q3 + 1.5 * IQR_col) %>%
  mutate(is_outlier = if_else(condition = daily_income > outliers_above,
                              true = TRUE, 
                              false = FALSE)) %>% 
  arrange(desc(daily_income), world_6region)
world_6region_vec <- distinct(billionaires_outliers, world_6region) %>% 
  arrange(world_6region) %>% 
  pull()  
outliers_above <- distinct(billionaires_outliers, world_6region, outliers_above) %>% 
  arrange(world_6region) %>% 
  pull() 

for (i in seq_along(world_6region_vec)) {
  all_oneregion <- billionaires_outliers %>% 
    filter(world_6region == world_6region_vec[i])
  outliers_oneregion <- all_oneregion %>% 
    filter(is_outlier == TRUE)
  y_axis_offset_for_outlier_label <- range(all_oneregion$daily_income) %>% 
    diff()
  y_axis_offset_for_outlier_label <- y_axis_offset_for_outlier_label * 0.05
  if (nrow(outliers_oneregion) == 0){
    cat(world_6region_vec[i], "has no outliers. I will plot all names.\n")
  p <- ggplot() + 
    geom_boxplot(data = all_oneregion,
                 mapping = aes(y = daily_income, x = 1),
                 color = "purple")   +
    geom_text_repel(mapping = aes(y = daily_income, 
                                  x = 1,
                                  label = person,
                            size = daily_income), 
                    data = all_oneregion,
                    max.overlaps = 100,  
                    force = 7,
                    alpha = 0.7,
                    segment.alpha = 0.2,
                    segment.size = 0.1) + 
    scale_x_continuous(breaks = NULL, 
                       name = world_6region_vec[i]) +
    scale_y_continuous(breaks = seq(from = 10^6, 
                                    to = ceiling(round(
                                    max(all_oneregion$daily_income) * 10^(-6)) * 10^6), 
                                    by = 10^7),
                       labels = as.character(
                                             seq(from = 10^6, 
                                                 to = ceiling(round(max(all_oneregion$daily_income) * 10^(-6)) * 10^6), 
                                                 by = 10^7)
                                   )
                       ) 
  ggsave(plot = p, filename = paste0("../my_output_files/outliers_billionaires_", world_6region_vec[i], ".pdf"),
         width = 7 * 2)
    
  } else {
  set.seed(155)
  p <- ggplot() + 
    geom_boxplot(data = all_oneregion,
                          mapping = aes(y = daily_income, x = 1),
                          color = "purple", outlier.size = 3)  +
    geom_text_repel(mapping = aes(y = daily_income, 
                                  x = 1,
                                  label = person,
                                  size = daily_income), 
                    data = outliers_oneregion,
                    max.overlaps = 100,  force = 7,
                    alpha = 0.7,
                    segment.alpha = 0.3,
                    segment.size = 0.2) + 
    scale_x_continuous(breaks = NULL, 
                       name = world_6region_vec[i]) +
    scale_size_continuous(range = c(6,10), 
                          breaks = seq(from = round(min(outliers_oneregion$daily_income)),
                                       to = round(max(outliers_oneregion$daily_income)),
                                       by = round(max(outliers_oneregion$daily_income) * 0.4))) +
    scale_y_continuous(breaks = seq(from = 10^6, 
                                    to = ceiling(round(
                                      max(all_oneregion$daily_income) * 10^(-6)) * 10^6), 
                                    by = 10^7),
                        labels = as.character(seq(from = 10^6, 
                                                  to = ceiling(round(
                                                    max(all_oneregion$daily_income) * 10^(-6)) * 10^6), 
                                                  by = 10^7))) +
    geom_hline(yintercept = outliers_above[i], 
               color = "seagreen", 
               linewidth = 1, 
               linetype=3, 
               alpha = 0.4) + 
    annotate(geom = "text", 
             x = 0.6, 
             y = outliers_above[i] + y_axis_offset_for_outlier_label, 
             label = outliers_above[i], 
             color = "seagreen", 
             size = 6) +
    theme(axis.text = element_text(size = 12), 
          axis.title = element_text(size = 14))
  cat(world_6region_vec[i], "\n") 
  print(p)
  ggsave(plot = p, filename = paste0(
    "../my_output_files/outliers_billionaires_", 
    world_6region_vec[i], ".pdf"),
         width = 7 * 2.2)
    
  }
}

```

