---
title: "Join tables with dplyr"
format: html
editor: visual
---

## Data and libraries

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(gapminder)
library(glue)

```

```{r}
marriage_df <- read_csv(file = glue("/lnet/aic/personal/cinkova/",
"R_BEGINNERS_SHORT/datasets_ATRIUM/gapminder_firstmarriage.csv"))
# or copy it to your location and read from there
str(marriage_df)
```

## Task

1.  Read the file as above and rename the column `2005` to "marriage". Retype this. Mind the quotes in "2005". Try without. They are very important when you get a column with an ambiguous name. With even weirder names you may even need backticks (\`2005\`).

    ```{r}

    ```

2.  Filter `gapminder` for year 2007.

3.  Join it with `marriage_df` (only matching countries).

4.  Plot linear models of the association between the age of the first marriage in women and `gdpPercap`, `pop`, and `lifeExp` (each in a separate plot).

5.  Plot the same three models but this time map continent on color.

Write a short interpretation of the models.

### Linear model of the association between the age of the first marriage in women and lifeExp

```{r}

```

### Linear model of the association between the age of the first marriage in women and gdpPercap

```{r}

```

### Linear model of the association between the age of the first marriage in women and pop

```{r}

```

### Linear model of the association between the age of the first marriage in women and lifeExp, by continents

### Linear model of the association between the age of the first marriage in women and gdpPercap, by continents

### Linear model of the association between the age of the first marriage in women and pop, by continents
